---
title: "Appendix 1: Code for '121 years of change: The birds of Nebraska National Forest'"
output: github_document
---

```{r,echo=F}
filepath <- "~/Dropbox/Manuscripts/sandhills_2023/"
```

```{r}
library(tidyverse)
library(lubridate)
library(terra)
library(data.table)
library(rnaturalearth)
# also requires rnaturalearthhires
```


```{r,eval=F}
gbif <- read.delim(paste0(filepath,"0104629-230530130749713.csv"),sep = "\t")

# remove unclassifiable
gbif$species[which(gbif$species=="Junco hyemalis"&
        gbif$infraspecificEpithet=="")] <- "Junco sp."

gbif$species[which(gbif$species=="Setophaga coronata"&
        gbif$infraspecificEpithet=="")] <- "Setophaga coronata/auduboni"

sp_list <-  c("Junco","hyemalis","Junco hyemalis",
             "Junco","oreganus","Junco oreganus",
             "Junco","aikeni","Junco aikeni",
             "Junco","mearnsi","Junco mearnsi",
             "Junco","cismontanus","Junco cismontanus",
             "Junco","caniceps","Junco caniceps",
             "Junco","shufeldti",
                    "Junco oreganus",
             "Junco","montanus",
                    "Junco oreganus",
             "Setophaga","coronata","Setophaga coronata",
             "Setophaga","","Setophaga coronata/auduboni")

renamer <- matrix(data=sp_list,
       nrow=(length(sp_list)/3),ncol=3,
       byrow = T) %>% as.data.frame()
colnames(renamer) <- c("genus","from","to")

for(i in 1:nrow(renamer)){
  index=which(gbif$infraspecificEpithet==renamer$from[i])
  gbif$species[index]=renamer$to[i]
}

# reduce to relevant columns
# all have coords??

gbif <- gbif%>%
  select(species, locality, individualCount,
         decimalLongitude, decimalLatitude, eventDate) %>%
  unique()

taxo_order <- read_csv(paste0(filepath,"sp_list_ne.csv"))
taxo_order <- taxo_order%>%
  rename(species = scientific)

gbif <- gbif%>%
  left_join(taxo_order,by="species")

missing <- gbif$species[which(is.na(gbif$order))]%>%unique()%>%
  as.data.frame
# write as file, then put new names
# write_csv(missing,paste0(filepath,"missing.csv"))

# rename missing taxa... there's several of them!

rename_taxon <- function(gbif,wrong,right,keep){
  index <- which(gbif$species==wrong)
  if(length(index)<1){
    return(gbif)
  }else{
    if(keep==F){
      gbif <- gbif[-index,]
    }else{
      gbif$species[index] <- right
    }
    return(gbif)
  }
}

missing <- read_csv(paste0(filepath,"missing.csv"))

for(i in 1:nrow(missing)){
  right <- missing$right[i]
  wrong <- missing$wrong[i]
  keep <- missing$keep[i]
  gbif <- rename_taxon(gbif=gbif,right=right,wrong=wrong,keep=keep)
}

# gbif <- gbif[-which(gbif$species==""),]
gbif$species[which(gbif$species%like%"villosus")] <- "Leuconotopicus villosus"

write_csv(gbif,paste0(filepath,"reduced_gbif.csv"))
```

We are also going to import and reformat eBird data. Note that for *Junco*, we are only using populations predesignated to subspecies since roughly 57% of western *Junco* are *hyemalis* in Nebraska; thus, we can't make broad assumptions. To contrast, 98% in eastern Nebraska are *hyemalis*. (Silcock and Jorgensen 2023)

```{r,eval=F}
# reformat eBird data
# add to ensure nothing is missing in GBIF!
# first run revealed some data lacking

ebird <- read.delim(paste0(filepath,
                           "ebd_US-NE_relJun-2023/ebd_US-NE_relJun-2023.txt"),
                    sep = "\t",quote="")

# remove unknowns
`%notlike%` <- Negate(`%like%`)
ebird <- ebird[which(ebird$CATEGORY%notlike%"slash"&
                       ebird$CATEGORY%notlike%"domestic"&
                       ebird$CATEGORY%notlike%"spuh"),]

# split some taxa for better resolution
# code has errors...
#ssp_rename <- function(from,to){
#  index <- which(ebird$SUBSPECIES.SCIENTIFIC.NAME==from)
#  if(length(index)<1){
#    print(paste0("No matches for ",from))
#  }else{
#    ebird$SCIENTIFIC.NAME[index] <- to
#    print(paste0(from," changed to ",to))
#  }
#}

# remove unclassifiable
sp_list <-  c("Setophaga coronata coronata","Setophaga coronata",
              "Setophaga coronata auduboni","Setophaga auduboni",
              "Junco hyemalis hyemalis/carolinensis","Junco hyemalis",
             "Junco hyemalis [oreganus Group]","Junco oreganus",
             "Junco hyemalis aikeni","Junco aikeni",
             "Junco hyemalis mearnsi","Junco mearnsi",
             "Junco hyemalis cismontanus","Junco cismontanus",
             "Junco hyemalis caniceps","Junco caniceps",
             "Junco hyemalis hyemalis/carolinensis/cismontanus",
                    "Junco hyemalis/cismontanus",
             "Junco hyemalis oreganus x mearnsi",
                    "Junco hyemalis oreganus x mearnsi",
             "Junco hyemalis mearnsi x aikeni",
                    "Junco hyemalis mearnsi x aikeni",
             "Setophaga coronata coronata","Setophaga coronata",
             "Setophaga coronata auduboni","Setophaga auduboni",
             "Setophaga coronata coronata x auduboni",
                    "Setophaga coronata coronata x auduboni")

renamer <- matrix(data=sp_list,
       nrow=(length(sp_list)/2),ncol=2,
       byrow = T) %>% as.data.frame()
colnames(renamer) <- c("from","to")

for(i in 1:nrow(renamer)){
  index=which(ebird$SUBSPECIES.SCIENTIFIC.NAME==renamer$from[i])
  ebird$SCIENTIFIC.NAME[index]=renamer$to[i]
}

# reduce to relevant columns
# all have coords??

ebird <- ebird[order(ebird$TAXONOMIC.ORDER),]

ebird <- ebird%>%
  select(SCIENTIFIC.NAME, OBSERVATION.COUNT,
         LONGITUDE, LATITUDE, OBSERVATION.DATE) %>%
  rename("species" = SCIENTIFIC.NAME,
         "individualCount" = OBSERVATION.COUNT, "decimalLongitude" = LONGITUDE,
         "decimalLatitude" = LATITUDE, "eventDate" = OBSERVATION.DATE)%>%
  unique()

ebird$species[which(ebird$species%like%"villosus")] <- "Leuconotopicus villosus"

write_csv(ebird,paste0(filepath,"reduced_ebird.csv"))
```

```{r,eval=F}
# note some longitudes are wrong; remove anything over 0
# locality columns causing problems for ebird

gbif <- read_csv(paste0(filepath,"reduced_gbif.csv"))%>%
  filter(decimalLongitude < 0) %>%
  select(-order,-english,-locality)

ebird <- read_csv(paste0(filepath,"reduced_ebird.csv"),quote="")

gbif <- rbind(ebird,gbif) %>%
  unique()

rm(ebird)
# load shapefile

nf <- vect(paste0(filepath,"S_USA.AdministrativeForest/S_USA.AdministrativeForest.shp"))

ne_nf <- nf$FORESTNAME[which(nf$FORESTNAME%like%"Nebraska")]

# select NE National Forest
ne_nf <- nf[which(nf$FORESTNAME==ne_nf)]

# get points in Nebraska NF
ne_crs <- crs(ne_nf)

gbif_vect <- vect(gbif,geom=c("decimalLongitude","decimalLatitude"),crs = ne_crs)

ne_gbif <- extract(y = gbif_vect,x = ne_nf)

# get non-empty rows
index <- which(!is.na(ne_gbif$FORESTNAME))

ne_nf_gbif <- gbif[index,]

write_csv(ne_nf_gbif,paste0(filepath,"ne_natl_forest_birds.csv"))
```
```{r,eval=F}
ne_nf_gbif <- read_csv(paste0(filepath,"ne_natl_forest_birds.csv"))

bessey <- which(ne_nf_gbif$decimalLatitude<42.5&ne_nf_gbif$decimalLongitude>-101)
mckelvie <- which(ne_nf_gbif$decimalLatitude>42.3&ne_nf_gbif$decimalLongitude>-102)
pine_ridge_oglala <- which(ne_nf_gbif$decimalLatitude>42.3&ne_nf_gbif$decimalLongitude<c(-102))

ne_nf_gbif$district <- "Unknown"

ne_nf_gbif$district[bessey] <- "Bessey"
ne_nf_gbif$district[mckelvie] <- "McKelvie"
ne_nf_gbif$district[pine_ridge_oglala] <- "Pine Ridge / Oglala"

write_csv(ne_nf_gbif,paste0(filepath,"ne_natl_forest_birds.csv"))
```

# District Summaries

```{r}
ne_nf_gbif <- read_csv(paste0(filepath,"ne_natl_forest_birds.csv"))

ne_nf_gbif$species <- as.factor(ne_nf_gbif$species)
ne_nf_gbif$eventDate <- as.Date(ne_nf_gbif$eventDate)
ne_nf_gbif$district <- as.factor(ne_nf_gbif$district)
```

## Bessey

```{r}
districtR <- function(ne_nf_gbif,district){
  sub.ne <- ne_nf_gbif[ne_nf_gbif$district==district,]
  sub.ne$eventDate <- year(sub.ne$eventDate)
  sp_list <- unique(sub.ne$species)
  hist(sub.ne$eventDate,main=district,breaks=100,
       xlab="Year")
  for(i in 1:length(sp_list)){
    sp_x <- sp_list[i]
    sp_ne <- sub.ne[which(sub.ne$species==sp_x),]%>%
      select(species,eventDate)
    if(nrow(sp_ne)<1){
      next
    }
    print(summary(sp_ne))
    hist(sp_ne$eventDate,main=sp_x,breaks=100,
       xlab="Year")
  }
}
```

```{r}
districtR(ne_nf_gbif = ne_nf_gbif,district = "Bessey")
```

## mckelvie

```{r}
districtR(ne_nf_gbif = ne_nf_gbif,district = "McKelvie")
```

## Pine Ridge / Oglala

```{r}
districtR(ne_nf_gbif = ne_nf_gbif,district="Pine Ridge / Oglala")
```

# Missing

What are the missing taxa indicated in this study?

```{r,eval=F}
# prep GBIF data
gbif <- read.delim(paste0(filepath,"0104629-230530130749713.csv"),sep = "\t")

# reduce to relevant columns
# all have coords??

gbif <- gbif%>%
  select(gbifID, institutionCode, catalogNumber, species, locality,
         decimalLongitude, decimalLatitude, year) %>%
  unique()

taxo_order <- read_csv(paste0(filepath,"sp_list_ne.csv"))
taxo_order <- taxo_order%>%
  rename(species = scientific)

# get missing taxa
missing <- read_csv(paste0(filepath,"missing.csv")) %>%
  filter(keep==F)%>%
  rename(species = wrong)

miss_tax <- gbif%>%
  inner_join(missing,"species")

miss_tax <- miss_tax[order(miss_tax$species),]

write_csv(miss_tax,paste0(filepath,"missing_list.csv"))
```

Many of the taxa that are listed as here but have not been confirmed in the state are exotics or fossil species. One collection appears to have placed all records from North America at the same locality (Musée des Confluences, Lyon) The following merit (brief) following up:

Species|Collection|Locality|Year|Note
-------|----------|--------|----|----
*Agelaius tricolor*|USGS|NA|1932|Mis ID?
*Dryobates scalaris*|Great Backyard Bird Count|69144 Keystone|2000|Mis ID?
*Larus heermanni*|USGS|NA|1933|Wrong locality?
*Larus occidentalis*|iNaturalist|NA|Valid record; outside study area
*Megascops kennicottii*|University of Michigan|Kearney|1925|Two birds; likely misidentified; checked with collection
*Passer montanus*|Cornell|Several||Multiple records; not accepted? Outside study area
*Picoides arcticus*|Yale Peabody Museum|Lincoln|1895|Transcription error; refers to Lincoln, Maine
*Poecile carolinensis*|Cornell|Omaha|2000|Not on state list; ouside study area
*Progne tapera*|USGS|NA|1923|No details; mis ID?
*Sialia mexicana*|USGS|Chadron|1922|Recently seen in state; predates current records?

Many thanks to Dr. Brett Benz, Dr. Ben Winger, and Dr. Kristof Zykowski for the specimen information.

I'm including *Sialia* in the study for now - hoping to confirm.

# Species through time

We can compare our lists through time to see how things have changed.

```{r,echo=F}
rm(list=ls())
filepath <- "~/Dropbox/Manuscripts/sandhills_2023/"
```

```{r}
zimmer <- read_csv(paste0(filepath,"zimmer_1913_species.csv")) %>%
  select(-Species) %>%
  rename("Zimmer_1913" = List)
ford <- read_csv(paste0(filepath,"ford_1957_species.csv")) %>%
  select(-Species)%>%
  rename("Ford_1957" = List)
bray <- read_csv(paste0(filepath,"Bray_Sandhills_Species.csv"))%>%
  rename("Bray_1993" = Bray) %>%
  select(-Species,-Notes)

ne_nf_gbif <- read_csv(paste0(filepath,"ne_natl_forest_birds.csv")) %>%
  filter(district == "Bessey") %>%
  select(species) %>%
  rename("SciName" = species) %>%
  unique()

ne_nf_gbif$GBIF_2023 <- "GBIF"
```

We have `r nrow(zimmer)` taxa from 1913, `r nrow(bray)` taxa, and `r nrow(ne_nf_gbif)` taxa from 2023.

```{r}
comp_years <- zimmer %>%
  full_join(ford,by = "SciName") %>%
  full_join(bray,by = "SciName") %>%
  full_join(ne_nf_gbif,by = "SciName")
```

```{r,eval=F}
# species recorded by GBIF
# debugging step
index <- which(!is.na(comp_years$GBIF_2023))

write_csv(comp_years[index,],paste0(filepath,"missing_gbif.csv"))
```

We can save this file as an appendix of species' occurrence.

```{r}
write_csv(comp_years,paste0(filepath,"compare_lists.csv"))
```

We can also compare all sites, and see what is in the other districts but not in Bessey yet.

```{r}
zimmer <- read_csv(paste0(filepath,"zimmer_1913_species.csv")) %>%
  select(-Species) %>%
  rename("Zimmer_1913" = List)
ford <- read_csv(paste0(filepath,"ford_1957_species.csv")) %>%
  select(-Species)%>%
  rename("Ford_1957" = List)
bray <- read_csv(paste0(filepath,"Bray_Sandhills_Species.csv"))%>%
  rename("Bray_1993" = Bray) %>%
  select(-Species,-Notes)

ne_nf_gbif <- read_csv(paste0(filepath,"ne_natl_forest_birds.csv")) %>%
  rename("SciName" = species)
# split by district
pr_og <- ne_nf_gbif %>% 
  select(SciName,district) %>%
  filter(district == "Pine Ridge / Oglala")%>%
  rename("Pine_Ridge" = district) %>%
  unique()

pr_og$Pine_Ridge[which(!is.na(pr_og$Pine_Ridge))] <- "Confirmed"

bessey <- ne_nf_gbif %>% 
  select(SciName,district) %>%
  filter(district == "Bessey")%>%
  rename("GBIF 2023" = district) %>%
  unique()

bessey2 <- ne_nf_gbif %>% 
  select(SciName,district) %>%
  filter(district == "Bessey")%>%
  rename("Bessey" = district) %>%
  unique()

mckelvie <- ne_nf_gbif %>% 
  select(SciName,district) %>%
  filter(district == "McKelvie")%>%
  rename("McKelvie" = district) %>%
  unique()

mckelvie$McKelvie[which(!is.na(mckelvie$McKelvie))] <- "Confirmed"

comp_years <- zimmer %>%
  full_join(ford,by = "SciName") %>%
  full_join(bray,by = "SciName") %>%
  full_join(bessey,by = "SciName") %>%
  full_join(bessey2,by = "SciName") %>%
  full_join(mckelvie,by = "SciName") %>%
  full_join(pr_og,by = "SciName")

# make Bessey the full list
comp_years$`GBIF 2023`[!is.na(comp_years$`GBIF 2023`)] <- "Confirmed"
comp_years$Bessey[which(comp_years$`GBIF 2023`=="Confirmed")] <- "Confirmed"
comp_years$Bessey[which(comp_years$Zimmer_1913=="Confirmed")] <- "Confirmed"
comp_years$Bessey[which(comp_years$Ford_1957=="Confirmed")] <- "Confirmed"
comp_years$Bessey[which(comp_years$Bray_1993=="Confirmed")] <- "Confirmed"
comp_years$Bessey[which(comp_years$Bray_1993=="Hypothetical"&is.na(comp_years$Bessey))] <- "Hypothetical"

# change bad scinames
comp_years$SciName[which(comp_years$SciName=="Empidonax occidentalis")] <- 
  "Empidonax difficilis"
comp_years$SciName[which(comp_years$SciName=="Coccothraustes vespertinus")] <- 
  "Hesperiphona vespertina"

scinums <- read_csv(paste0(filepath,"scinums.csv")) %>%
  rename("SciName" = `Scientific Name`)

comp_years_full <- comp_years %>%
  left_join(scinums,by = "SciName")

# replace British names
comp_years_full$SciName <- gsub(pattern = "grey",replacement = "gray",
                                x = comp_years_full$SciName)
comp_years_full$SciName <- gsub(pattern = "Grey",replacement = "Gray",
                                x = comp_years_full$SciName)
# other changes to naming done in spreadsheet

comp_years_full <- comp_years_full[order(comp_years_full$IOC),]
comp_years_full <- comp_years_full %>%
  select(IOC,`English name`,SciName,Zimmer_1913,Ford_1957,Bray_1993,
         `GBIF 2023`,Bessey,McKelvie,Pine_Ridge) %>%
  rename("Scientific name" = SciName,
         "Zimmer 1913" = Zimmer_1913,
         "Ford 1957" = Ford_1957,
         "Bray 1993" = Bray_1993,
         "Pine Ridge / Oglala" = Pine_Ridge)

write_csv(comp_years_full,paste0(filepath,"all_sites_compare.csv"))
```

# Figures

Here, I document the creation of figures for the manuscript.

```{r}
nf <- vect(paste0(filepath,"S_USA.AdministrativeForest/S_USA.AdministrativeForest.shp"))
ne_nf <- nf$FORESTNAME[which(nf$FORESTNAME%like%"Nebraska")]
ne_nf <- nf[which(nf$FORESTNAME==ne_nf)]
ne_crs <- crs(ne_nf)

# get a map of Nebraska Counties
county <- ne_download(scale = "large",type = "admin_2_counties",
                      returnclass = "sf") %>% vect() %>% project(ne_crs)

# get Nebraska
county_ne <- county[county$ISO_3166_2=="US-31"]

# crop forest to Nebraska only for map
# small sliver polygon at border, but ok
ne_nf2 <- crop(ne_nf,county_ne)

# county shapefile 

plot(ne_nf2,xlab="Longitude",ylab="Latitude")
plot(county_ne,add=T)
plot(ne_nf2,col="gray",add=T)
for(i in 1:length(county_ne)){
  cr <- county_ne[i]
  center <- centroids(cr)
  name <- cr$NAME
  if(name=="Morrill"|name=="Garden"){
    lab <- county_ne[which(county_ne$NAME=="Scotts Bluff")]%>%
      centroids()%>%
      geom()%>%
      as.data.frame()
    center1 <- center %>% geom() %>% as.data.frame()
    lab$x <- center1$x
    center <- vect(lab,geom=c("x","y"),crs=ne_crs)
  }
  text(center,labels=name,halo=T)
}

# state map to compare

plot(county_ne,col="lightgray")
plot(ne_nf2,col="black",add=T)
```

```{r}
# forest size
# divide up by county

# pine ridge oglala in Dawes, Sioux
# mckelvie in Cherry
# bessy in Thomas, Blaine

count_cropper <- function(counties){
  cr <- county[contains(counties,vars = county$NAME)]
  cr_crop <- crop(ne_nf2,cr)
  plot(cr_crop)
  return(expanse(cr_crop,unit="km"))
}

# area of Pine Ridge
count_cropper(counties = c("Dawes","Sioux"))
# area of McKelvie
count_cropper(counties = c("Cherry"))
# area of Bessey
count_cropper(counties = c("Thomas","Blaine"))
```

Check the locality of the Prairie Warbler.

```{r}
praw <- c(41.905982,-100.319638)

plot(ne_nf2,xlim=c(-100.31960,-100.31964),
     ylim=c(41.90595,41.90599))
points(y=praw[1],x=praw[2],pch=20)
```

```{r}
praw <- as.data.frame(praw)
praw_coord <- matrix(data = c(-100.31964,41.90598),
                     nrow = 1,ncol=2) %>%
  vect(type="points",atts=NULL,crs=crs(ne_nf2))

# distance in meters
# function seems to be malfunctioning? m and km give same output...

terra::distance(x=ne_nf2,
         y=praw_coord,unit="m")
```

